|

### 2024-09-03 manuscript/paperCMSACversion.Rmd
small edits

Fighting in the National Hockey League (NHL) has been an integral part of the sport since the earliest days of the league. More than just a way to express frustration, players often use fighting strategically to the benefit of the team. One of these strategies includes instigating a fight when a player from the opposite team has violated an unwritten rule of the game, which historically has led to hockey teams carrying so called "enforcers" on their rosters. An enforcer is a member of a hockey team who, along with being a regular player, is the one tasked with fighting these opposing players. This is not an officially defined position but the nature of needing to be able to fight any opponent generally designates the players most likely to be enforcers.

@Colburn1986 discusses the intentionality behind these public fights. The crowd gets energized and entertained, but more importantly, the fight is a stage used to establish the social customs and expectations of players on the ice.  Specifically, they say, "insofar as the fist-fight is the locus of socially available beliefs concerning the moral character and integrity of players, ... the fist-fight represents a mode of informal social control among players" .

@@

Recently, @Popkin2023 explored the effects of being an enforcer on mortality rates, ages at death, and causes of death. They defined enforcers in two ways: (1) those who had 50 or more career fights (E-F cohort) and (2) those who had, on average, more than 3 penalty minutes per game (E-P cohort).

Based on their analysis (a matched cohort study) they find that there is no statistically significant difference in the mortality rates of between enforcers and controls for either definition of enforcer, but they do conclude that enforcers died on average about 10 years earlier than their non-enforcer counterparts.  In addition, they go on to draw the conclusion that there is not a significant difference in all-cause mortality rate but find enforcers died more frequently of drug overdose and suicide.

However, we believe that some of the methods used in @Popkin2023 to reach these conclusions are not statistically appropriates.  Most notably, the method used to reached their most sensational finding - enforcers died on average about 10 years earlier than their non-enforcer counterparts - is statistically unsuitable.  Specifically, to reach this conclusion they use the data in the matched cohort study, but then condition on death.  This effectively breaks the matching that they appropriately used to compare mortality rates, and therefor this conclusion is not statistically valid.

Here, we attempt to replicate the results of the analysis in @Popkin2023, but what we believe to be more appropriate statistical analysis, specifically survival analysis methodology.  This shift in perspective allows us to overcome the weakness of mortality rate, which is that it is a function of when it is calculated (i.e. Eventually, the mortality rate will always be 100% when all members of the original sample have died). Instead, survival analysis provides a survival curve, which can be used to analyze how the probability of survival changes over time.  In addition, we augmented the data used in @Popkin2023 to increase the sample size in the analysis and include more deceased players in the data set.

@@

Figure \@ref(fig:estsurv) shoes the estimated survival curves for enforcers versus non-enforcers (E-P) for defenders with all continuous variables held constant at mean levels.  Using this, the estimated survival rate at the age of 50 for enforcers vs non-enforcers is 96.57\% (95\% CI: 94.63\%, 98.56\%) vs 98.49\% (95\% CI: 97.98\%, 99.00\%).  For the same conditions at the age of 75, the estimated survival rate for enforcers is about 74.4\% (95\% CI: 61.4\%, 90.32\%) compared to non-enforcers estimate survival rate of 87.95\% (95\% CI: 83.09\%, 93.10\%).

@@

```{r fig:estsurv, echo = FALSE, fig.cap = "Estimated Survival curve for EP for defenders.  All other variables are fixed at their mean level (GP = 285.8, Ht = 72.7in, Wt = 197.03, and Year of birth 1974.5."}
coxphmod_EP <- coxph(Surv ~ EP + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
# summary(coxphmod_EP)
# library(xtable)
# xtable(summary(coxphmod_EP)$coef[,-4],digits = 5)
# cox.zph(coxphmod_EP)

>>>>

Fighting in the National Hockey League (NHL) has been an integral part of the sport since the earliest days of the league. More than just a way to express frustration, players often use fighting strategically to the benefit of the team. One of these strategies includes instigating a fight when a player from the opposite team has violated an unwritten rule of the game, which historically has led to hockey teams carrying so-called "enforcers" on their rosters. An enforcer is a member of a hockey team who, along with being a regular player, is the one tasked with fighting these opposing players. While this is not an officially defined position, the nature of needing to be able to fight any opponent generally designates the biggest and strongest players as those most likely to be enforcers.

@Colburn1986 discusses the intentions behind these public fights. The crowd gets energized and entertained, but more importantly, the fight is a stage used to establish the social customs and expectations of players on the ice.  Specifically, they say, "insofar as the fist-fight is the locus of socially available beliefs concerning the moral character and integrity of players, ... the fist-fight represents a mode of informal social control among players".

@@

Recently, @Popkin2023 explored the effects of being an enforcer on mortality rates, ages at death, and causes of death to address the long-term impact this repeated conflict might have. They defined enforcers in two ways: (1) those who had 50 or more career fights (E-F cohort) and (2) those who had, on average, more than 3 penalty minutes per game (E-P cohort).

Based on their analysis (a matched cohort study) they find that there is no statistically significant difference in the mortality rates between enforcers and controls for either definition of enforcer, but they do conclude that enforcers died on average about 10 years earlier than their non-enforcer counterparts. In addition, they draw the conclusion that there is not a significant difference in all-cause mortality rate but find enforcers died more frequently of drug overdose and suicide.

However, we believe that some of the methods used in @Popkin2023 to reach these conclusions are not statistically appropriates.  Most notably, the method used to reached their most sensational finding - enforcers died on average about 10 years earlier than their non-enforcer counterparts - is statistically unsuitable. Specifically, to reach this conclusion, they use the data from the matched cohort study but then condition on death.  This effectively breaks the matching that they appropriately used to compare mortality rates, and therefore this conclusion is not statistically valid.

Here, we attempt not to replicate the results of the analysis in @Popkin2023, but what we believe to be more appropriate statistical analysis, specifically survival analysis methodology. This shift in perspective allows us to overcome the weakness of mortality rate, which is that it is a function of when it is calculated (i.e. In all cases, mortality rate will be 100% when all members of the original sample have died). Instead, survival analysis provides a survival curve, which can be used to analyze how the probability of survival changes over time.  In addition, we augmented the data used in @Popkin2023 to increase the sample size in the analysis and include more deceased players in the data set.

@@

Figure \@ref{fig:estsurv} shoes the estimated survival curves for enforcers versus non-enforcers (E-P) for defenders with all continuous variables held constant at mean levels.  Using this, the estimated survival rate at the age of 50 for enforcers vs non-enforcers is 96.57\% (95\% CI: 94.63\%, 98.56\%) vs 98.49\% (95\% CI: 97.98\%, 99.00\%).  For the same conditions at the age of 75, the estimated survival rate for enforcers is about 74.4\% (95\% CI: 61.4\%, 90.32\%) compared to non-enforcers estimate survival rate of 87.95\% (95\% CI: 83.09\%, 93.10\%).

@@

```{r estsurv, echo = FALSE, fig.cap = "Estimated Survival curve for EP for defenders.  All other variables are fixed at their mean level (GP = 285.8, Ht = 72.7in, Wt = 197.03, and Year of birth 1974.5."}
coxphmod_EP <- coxph(Surv ~ EP + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
# summary(coxphmod_EP)
# library(xtable)
# xtable(summary(coxphmod_EP)$coef[,-4],digits = 5)
# cox.zph(coxphmod_EP)
